Making ice sheet models \\ scale properly

### 2023-01-02 talk/slides.tex
progress on slides

\newcommand{\ba}{\mathbf{a}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bc}{\mathbf{c}}
\newcommand{\bbf}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bw}{\mathbf{w}}

\newcommand{\bF}{\mathbf{F}}
\newcommand{\bG}{\mathbf{G}}
\newcommand{\bQ}{\mathbf{Q}}

\newcommand{\grad}{\nabla}
\newcommand{\Div}{\nabla\cdot}

\newcommand{\argmin}{\operatorname{argmin}}

\newcommand{\CC}{\mathbb{C}}
\newcommand{\EE}{\mathbb{E}}
\newcommand{\RR}{\mathbb{R}}

\newcommand{\ddt}[1]{\ensuremath{\frac{\partial #1}{\partial t}}}
\newcommand{\ddx}[1]{\ensuremath{\frac{\partial #1}{\partial x}}}
\newcommand{\Matlab}{\textsc{Matlab}\xspace}
\newcommand{\Octave}{\textsc{Octave}\xspace}
\newcommand{\eps}{\epsilon}

\newcommand{\ip}[2]{\left<#1,#2\right>}

\newcommand{\xiphalf}{{x_{i+\frac{1}{2}}}}
\newcommand{\ximhalf}{{x_{i-\frac{1}{2}}}}
\newcommand{\Fiphalf}{{F_{i+\frac{1}{2}}}}
\newcommand{\Fimhalf}{{F_{i-\frac{1}{2}}}}
\newcommand{\Fiphalfn}{{F^n_{i+\frac{1}{2}}}}
\newcommand{\Fimhalfn}{{F^n_{i-\frac{1}{2}}}}

\newcommand{\trefcolumn}[1]{\begin{bmatrix} \phantom{x} \\ #1 \\ \phantom{x} \end{bmatrix}}
\newcommand{\trefmatrixtwo}[2]{\left[\begin{array}{c|c|c} & & \\ #1 & \dots & #2 \\ & & \end{array}\right]}
\newcommand{\trefmatrixthree}[3]{\left[\begin{array}{c|c|c|c} & & & \\ #1 & #2 & \dots & #3 \\ & & & \end{array}\right]}
\newcommand{\trefmatrixgroups}[4]{\left[\begin{array}{c|c|c|c|c|c} & & & & & \\ #1 & \dots & #2 & #3 & \dots & #4 \\ & & & & & \end{array}\right]}

\newcommand{\blocktwo}[4]{\left[\begin{array}{c|c} #1 & #2 \\ \hline #3 & #4 \end{array}\right]}

\newcommand{\bqed}{{\color{blue}\qed}}
\newcommand{\ds}{\displaystyle}

\newcommand\mynum[1]{{\renewcommand{\insertenumlabel}{#1}%
      \usebeamertemplate{enumerate item} \,}}

@@

\begin{frame}
  \maketitle
\end{frame}

@@

\begin{itemize}
\item before numerical models \dots \emph{what} are we modeling?
\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width ratio
\item<5> it is \emph{not} sea ice
\end{itemize}

\bigskip
\begin{minipage}[t][40cm][t]{\textwidth}
\begin{center}
\includegraphics<1>[height=0.64\textheight]{images/ant-pittard2021.png}
\only<1>{\par {\scriptsize Antarctic ice sheet (Pittard et al 2021)}}
\only<2>{\vspace{15mm}}
\includegraphics<2>[width=\textwidth]{images/ant-schoofhewitt2013.png}
\only<2>{\vspace{10mm}}
\only<2>{\par {\scriptsize note vertical exageration and rough bed (Schoof \& Hewitt 2013)}}
\includegraphics<3>[height=0.64\textheight]{images/alps-seguinot2018.png}
\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum (Seguinot et al 2018)}}
\includegraphics<4>[height=0.64\textheight]{images/british-clark2022.png}
\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum (Clark et al 2022)}}
\includegraphics<5>[height=0.70\textheight]{images/not-sea-ice.png}
\end{center}
\end{minipage}
\end{frame}

@@

@@

\newcommand{\oo}[1]{\displaystyle O\left(#1\right)}

@@

@@

\begin{itemize}
\footnotesize
\item E.~Hazan, A.~Agarwal, \& S.~Kale (2007).  \href{https://link.springer.com/content/pdf/10.1007/s10994-007-5016-8.pdf}{\emph{Logarithmic regret algorithms for online convex optimization.}} Machine Learning, 69(2), 169-192
    \begin{itemize}
    \scriptsize
    \item[$-$] $O(\log j)$ regret bounds for positive definite Hessians, and for Newton algorithms
    \end{itemize}
\item D.~P.~Kingma \& J.~Ba (2014). \href{https://arxiv.org/abs/1412.6980}{\emph{Adam: A method for stochastic optimization}}, preprint arXiv:1412.6980.
    \begin{itemize}
    \scriptsize
    \item[$-$] claims $O(\sqrt{j})$ regret bound
    \end{itemize}
\item M.~Zinkevich (2003). \href{https://www.aaai.org/Papers/ICML/2003/ICML03-120.pdf}{\emph{Online convex programming and generalized infinitesimal gradient ascent}}, Proceedings of the 20th International Conference on Machine Learning, 928-936
    \begin{itemize}
    \scriptsize
    \item[$-$] introduced regret
    \item[$-$] $O(\sqrt{j})$ regret bound of OGD
    \end{itemize}
\end{itemize}
\end{frame}

>>>>

\newcommand{\eps}{\epsilon}
\newcommand{\RR}{\mathbb{R}}

\newcommand{\grad}{\nabla}
\newcommand{\Div}{\nabla\cdot}
\newcommand{\trace}{\operatorname{tr}}

\newcommand{\hbn}{\hat{\mathbf{n}}}

\newcommand{\bb}{\mathbf{b}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bbf}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}

\newcommand{\bF}{\mathbf{F}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bX}{\mathbf{X}}

\newcommand{\bxi}{\bm{\xi}}

\newcommand{\bzero}{\bm{0}}

\newcommand{\rhoi}{\rho_{\text{i}}}

\newcommand{\ip}[2]{\left(#1,#2\right)}

\newcommand{\mR}{R^{\bm{\oplus}}}
\newcommand{\iR}{R^{\bullet}}

\newcommand{\nn}{{\text{n}}}
\newcommand{\pp}{{\text{p}}}
\newcommand{\qq}{{\text{q}}}
\newcommand{\rr}{{\text{r}}}

\newcommand{\bus}{\bu|_s}
\newcommand{\oo}[1]{\displaystyle O\left(#1\right)}
\newcommand{\sold}{s_{\text{o}}}

@@

%\begin{frame}
%  \maketitle
%\end{frame}

{
  \usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/gray-british-clark2022.png}}
  \begin{frame}
    \titlepage
  \end{frame}
}

@@

\begin{itemize}
\item before numerical modeling \dots \emph{what} are we modeling?
\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width ratio
\end{itemize}

\bigskip
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\includegraphics<1>[height=0.69\textheight]{images/ant-pittard2021.png}
\only<1>{\par {\scriptsize Antarctic ice sheet}} % (Pittard et al 2021)
\only<2>{\vspace{17mm}}
\includegraphics<2>[width=\textwidth]{images/ant-schoofhewitt2013.png}
\only<2>{\vspace{13mm}}
\only<2>{\par {\scriptsize note vertical exaggeration and rough bed (Schoof \& Hewitt 2013)}}
\includegraphics<3>[height=0.69\textheight]{images/alps-seguinot2018.png}
\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum (Seguinot et al 2018)}}
\includegraphics<4>[height=0.69\textheight]{images/british-clark2022.png}
\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum (Clark et al 2022)}}
\includegraphics<5>[height=0.69\textheight]{images/not-sea-ice.png}
\only<5>{\par {\scriptsize but an ice sheet is \emph{not} sea ice!}}
\end{center}
\end{minipage}
\end{frame}

\begin{frame}{basic facts}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon!
    \end{itemize}
\item ice sheet geometry and velocity evolve \emph{in contact with the climate}
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] melt and calving when floating in the ocean (\emph{ice shelves})
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming performance analysis, I will be ignoring quite a bit of detail!
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (calving, crevasses)
    \end{itemize}
\item now, a simple definition of \emph{ice sheet model}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}
\end{column}
\end{columns}
\end{frame}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
an \alert{ice sheet model} is a map
\end{definition} 
\item two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$a(t,x) = \begin{pmatrix}
\text{balance of solid precipitation} \\
\text{\, vs melt \& runoff at surfaces}
\end{pmatrix}$$

\vspace{-3mm}
        \begin{itemize}
        \item[{\scriptsize $\bullet$}] units of mass flux:\, $\text{kg}\, \text{m}^{-2} \text{s}^{-1}$
        \end{itemize}

    \item[$\circ$] \emph{bed elevation} $b(x)$ [m]
    \end{itemize}
\item one output:
    \begin{itemize}
    \item[$\circ$] \emph{upper surface elevation} $s(t,x)$ [m]
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}

\vspace{2mm}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{conservation}

\begin{itemize}
\item but this is a mathematics talk in fact
\item FIXME conservation laws; conserve momentum standard; conserve mass means both conserve mass in interior (incompressible) and at surface (SKE)
\end{itemize}
\end{frame}

\begin{frame}{strong form of ice sheet model}

\begin{itemize}
\item FIXME RETURN TIME DERIV
\item nonlinear complementarity problem coupled to Stokes:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega$} \\
s - \Delta t\, \bu|_s \cdot \bn_s - (\sold + \Delta t\, a )&\ge 0 && \text{''} \\
(s - b) (s - \Delta t\, \bu|_s \cdot \bn_s - (\sold + \Delta t\, a)) &= 0 && \text{''} \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\bu &= \bzero && \text{on $\Gamma_0$} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$}
\end{align*}

    \begin{itemize}
    \item with regularized Glen-law effective viscosity ($\text{p}=\frac{1}{\nn}+1$):
      $$\nu_\eps(D\bu) = \frac{\Gamma}{2} \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
    \end{itemize}
\end{itemize}
\end{frame}

\section{time-stepping and stability}

@@

@@

\section{Stokes solver scaling}

\section{scalable implicit steps?}

@@

\begin{itemize}
\footnotesize
\item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
\item x
\end{itemize}
\end{frame}
